---
title: "Access to Basic Sanitation in Africa"
author: "Niall Redmond"
date: "2025-04-25"
format: html
theme: lumen
backgroundcolor: white
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

#Libraries
library(tidyverse)
library(plotly)
library(dplyr)

#data

unicef_indicator_1 <- read_csv("unicef_indicator_1.csv")
unicef_metadata_all <- read_csv("unicef_metadata.csv")
countries_continents <- read_csv("Countries-Continents.csv")

#transformations

unicef_metadata <-  filter(unicef_metadata_all, year >= 2000) #We only want years after 1999 so the data matches

colnames(unicef_metadata)[10] <- "life_expectancy" #Changing the name to something a little easier to use
colnames(unicef_metadata)[6] <- "population"
colnames(unicef_indicator_1)[6] <- "year" #make the names in both tables the same

data_join <- full_join(unicef_indicator_1, unicef_metadata)


```
###### This report examines the status of acess to **at least basic sanitation** and **life expectancy** in *Africa*.
```{r}

map_world <- map_data("world")

unicef_indicator_1_2021 <-  unicef_indicator_1 %>%
  filter(year == 2021)

map_data_join_2021 <- full_join(unicef_indicator_1_2021 , map_world, by = c("country" = "region"))

ggplot(map_data_join_2021) +
  aes(x = long, y = lat, group = group, fill = obs_value) +
  geom_polygon()+
  scale_fill_gradient(high="green",low = "red", na.value = "grey")+
  labs(fill = "%",
     title = "Global Access to Basic Sanitation in 2021",
     x = "Longitude",
     y = "Latitude"
     ) +
    theme_classic()




```

###### *Africa* stands out as being notably behind the world average in access to at least basic sanitation. The maps below compare access to basic sanitation and life expectancy.

::::: columns
::: column

```{r}

data_join_continents <- full_join(data_join, countries_continents, by = c("country" = "Country"))

DJ_Africa <- filter(data_join_continents , Continent == "Africa")


DJ_Africa_2021 <-  DJ_Africa %>%
  filter(year == 2021)

map_africa_data_join_2021 <- full_join(DJ_Africa_2021 , map_world, by = c("country" = "region"))

ggplot(map_africa_data_join_2021) +
  aes(x = long, y = lat, group = group, fill = obs_value) +
  geom_polygon()+
  scale_fill_gradient(high="green",low = "red", na.value = "grey")+
  labs(fill = "%",
       title = "Access to Basic Sanitation in Africa 2021",
       x = "Longitude",
       y = "Latitude"
  ) +
  theme_classic()+
  coord_cartesian(xlim = c(-20, 60), ylim = c(-40, 40))  # Zoom in on Africa

```
:::
::: column

```{r}

ggplot(map_africa_data_join_2021) +
  aes(x = long, y = lat, group = group, fill = life_expectancy) +
  geom_polygon()+
  scale_fill_gradient(high="green",low = "red", na.value = "grey")+
  labs(fill = "%",
       title = "Life Expectancy in Africa 2021",
       x = "Longitude",
       y = "Latitude"
  ) +
  theme_classic()+
  coord_cartesian(xlim = c(-20, 60), ylim = c(-40, 40))  # Zoom in on Africa
```
:::
:::::
```{r}

timeseries_plot_1 <- data_join_continents %>%
  filter(!is.na(Continent)) %>%  ggplot() +
  aes(year, obs_value, group = country, color = Continent )+
  geom_line(size=1.3, alpha = .8)+
  labs(
    x = "Year",
    y = "Access to Basic Sanitation",
    title = "Access to Basic Sanitation in the World"
  )+
  theme_bw()


ggplotly(timeseries_plot_1)

```
###### We can see *the peach lines of Africa* gathering at the bottom of the scatter plot above. For life expectancy, the bar chart below tells a similar story to access to basic sanitation.
```{r}
data_join_continents %>%
  group_by(Continent, year) %>%
  summarise(m_lifeexp = mean(life_expectancy, na.rm = TRUE)) %>%
    
    ggplot()+
    aes(Continent, m_lifeexp, fill = Continent) +
    geom_col()+
    labs(
    x = "Continent",
    y = "Life Expectancy",
    title = "Life Excpectancy by Continent Bar Chart"
  )+
  theme_light()



```


::::: columns
::: column


```{r}
DJ_ethiopia <- filter(data_join, country == "Ethiopia")

timeseries_plot_2 <- DJ_ethiopia %>%
  ggplot() +
  aes(year, obs_value, color = country )+
  geom_line()+
  labs(
    x = "Year",
    y = "Access to Basic Sanitation",
    title = "Access to Basic Sanitation in Ethiopia"
  )+
  guides(color = "none", country = "none")+
  theme_light()


ggplotly(timeseries_plot_2)

```
:::
::: column


```{r}

timeseries_plot_3 <- DJ_ethiopia %>%
  ggplot() +
  aes(year, life_expectancy, color = country )+
  geom_line()+
  labs(
    x = "Year",
    y = "Life Expectancy",
    title = "Life Expectancy in Ethiopia"
  )+
  guides(color = "none", country = "none")+
  theme_light()

ggplotly(timeseries_plot_3)
```
:::
:::::


###### One could intuit that there is a strong positive correlation between life expectancy and access to basic sanitation. Focusing on Ethiopia above seems to confirm that intuition. Below, the scatter plot shows countries in Africa with over 10 million people. **The general trend towards the top right indicates the positive correlation between life expectancy and access to basic sanitation.**



```{r}
average_pop <- DJ_Africa %>%  #I want to only have the largest countries in africa,
  group_by(country) %>%
  mutate(avg_population = mean(population)) #so I make a new data frame with average population appended. 
                                            #average population is meaningless for data analatics, Im just using it temporarily

largest_countries <- average_pop %>%  
  filter(avg_population > 10000000) #I filter out the small countries


ggplot(largest_countries ) +
  aes(obs_value, life_expectancy, color = country, group = country, year = 2000) +
  geom_point()+
  labs(
    x = "Access to Basic Sanitation",
    y = "Life Expectancy",
    title = "Access to Basic Sanitation and Life Expectancy Scatter Plot"
  )+
  guides(color = "none", country = "none")+
  theme_light()


```

##### **In conclusion, the data suggests that a critical factor in increasing life expectancy is to increase access to at least basic sanitation.**
